\documentclass[12pt]{article}
% \usepackage{palatino}
\usepackage{epsfig}
\usepackage{epstopdf} % for \psfig{.eps}
\usepackage{amssymb} % for \varnothing
\usepackage{amsmath} % for \pmod
\usepackage{url} % to split long names
\usepackage{color}
\usepackage{listings}
\usepackage{hyperref}
\setlength{\topmargin}{-0.1in}
\setlength{\textheight}{8.0in}
% \newcommand\note[1]{{\color{red} #1}}
\pagestyle{myheadings}
\markboth{}{An Interpreter for a Simple Programming Language}
\newcommand\note[1]{\raggedright\fbox{#1}}
% \newcommand\ind{\hspace*{.3in}}
\newcommand\code[1]{\texttt{\textbf{#1}}}
\def\>{\hspace*{0.2in}}
\newenvironment{notelist}{\begin{list}
   {$\bullet$}
   {\setlength{\itemsep}{0in}}}
   {\end{list}}

\definecolor{light-gray}{gray}{0.75}

% Options for code listings
\lstset{
    aboveskip=\baselineskip,
    basicstyle=\ttfamily\small,
    frame=none,
    language=Haskell,
    numberstyle=\ttfamily\small\color{light-gray},
    numbers=right,
    showstringspaces=false,
}

\setlength{\parindent}{0pt}
\setlength{\parskip}{\baselineskip}

\begin{document}

\thispagestyle{empty}

\begingroup
    \centering
    \LARGE \textbf{Programming Assignment} \par
    \large \textbf{An Interpreter for a Simple Programming Language} \par
\endgroup

In this programming assignment, you will implement an interpreter for a simple programming language we will call LANG.
You will need to write a parser that takes a LANG program (as a \code{String}) and returns a parse tree for the program.
Then, you will write the interpreter that executes that parse tree.

Included with this assignment are two Haskell source files: 

\begin{notelist}
    \item \code{Parsing.hs} is a module that defines the basic monadic parsing API we studied in the section on
          monad transformers. You will use these parsing primitives to implement a recursive-descent parser for
          LANG programs.
    \item \code{AST.hs} is a module that defines the data types representing the abstract syntax of a LANG program.
          Your parser should use these types as its output and your interpreter should for its input.
\end{notelist}

\section{The Language}

The structure of a LANG program is reflected in the types defined in \code{AST.hs}:

\begin{notelist}
    \item A program is a list of statements.
    \item A statement is one of the following:
        \begin{notelist}
            \item A print statement for printing numeric values: \code{print (1 + 1)}
            \item A print statement for string literals: \code{sprint "Hello world!"}
            \item A read statement to read a number from the console and store that value in a variable: \code{read x}
            \item An assignment statement: \code{x = 42}
            \item A while loop that executes a statement for as long as a control expression evaluates to a true value:
\begin{lstlisting}
    while x < 10 {
        x = x + 1
        print x
    }
\end{lstlisting}
            \item An if-then statement that executes a statement if a control expression evaluates to a true value:
\begin{lstlisting}
    if x < 10 then
        sprint "Less than ten"
\end{lstlisting}
            \item An if-then-else statement that executes one statement if a control expression evaluates to a true value but
                  executes a second statement otherwise:
\begin{lstlisting}
  if x < 10 then
      sprint "Less than ten"
  else
      sprint "Greater than or equal to ten"
\end{lstlisting}
            \item A compound statement, surrounded by braces, that can contain zero or more other statements. Since the abstract syntax
                  for \code{if} and \code{while} statements only admits a single statement body, compound statements allow for more
                  complex constructs. A compound statement is seen in the sample \code{while} loop above.
        \end{notelist}
    \item An expression is one of the following:
        \begin{notelist}
            \item A variable reference that evaluates to the variables current value in the global name scope.
            \item A literal integer constant. You only need to support decimal notation with optional negation using \code{~} (e.g., \code{~6} for negative six).
            \item Binary operator application. See the description of binary operators below.
            \item Unary operator application. The only unary operator you need to support is \code{!} for logical negation.
        \end{notelist}
    \item \code{read} statements, assignment statements, and variable references require the notion of identifiers. Your
          parser should support identifiers that start with an alphabetic character followed by zero or more alphanumeric
          characters.
    \item \code{sprint} uses string literals. Your parser does not need to support escape sequences in string literals.
\end{notelist}

\subsection{Expressions, Arithmetic, and Logic}

\begin{notelist}
    \item LANG does not support binary operators with different precedence or associativity. As such, binary expressions must
          be surrounded with parentheses if they are subexpression of a larger expression.
    \item All arithmetic in LANG is with integers, specifically Haskell's arbitrary-precision integers. The division operator
          should perform integer division only.
    \item Relational and comparison operators should return \code{1} for true and \code{0} for false.
    \item Logical operators (including unary logical negation) should treat non-zero values as true and \code{0} as false.
\end{notelist}

The following binary operators should be supported:

\begin{notelist}
    \item \code{+} -- Addition
    \item \code{-} -- Subtraction
    \item \code{*} -- Multiplication
    \item \code{/} -- Division
    \item \code{\^} -- Exponentiation
    \item \code{\%} -- Modulus
    \item \code{==} -- Equality
    \item \code{>} -- Comparison: greater than
    \item \code{<} -- Comparison: less than
    \item \code{>=} -- Comparison: greater than or equal to
    \item \code{<=} -- Comparison: less than or equal to
    \item \code{\&\&} -- Logical AND
    \item \code{||} -- Logical OR
\end{notelist}

\section{Implementation}

Copy the starter files (\code{Parsing.hs} and \code{AST.hs}) into a directory for your implementation.
Create a Haskell source file \code{Main.hs} with a function \code{main :: IO ()} to hold your 
program's main entry point.

Create separate modules for your parser and interpreter and import those into your \code{Main.hs}. 
The parsing module will need to import \code{Parsing}, the module defined in \code{Parsing.hs} and
the \code{AST} module in \code{AST.hs}. Your interpreter module will just need the \code{AST} module.

You main function should use the \code{System.Environment} module to read your program's command-line
arguments. If exactly one argument was given, treat that as the file name of the source file to
interpret. Load the the contents of that file as a string and parse it. If the parse was successful,
interpret the resulting \code{Program} value. If parse was not successful, report the message to the 
user and exit.

You should be able to compile your program into an executable with:

\begin{lstlisting}
> ghc Main.hs
\end{lstlisting}

You should also be able to run your program without compiling it with:

\begin{lstlisting}
> runghc Main.hs
\end{lstlisting}

\subsection{Implementing the parser}

The \code{Parsing} module defines the \code{Parser} type, a state monad with the \code{ExceptT}
transformer adding the possibility of failure with \code{ParseError} values. Your ultimate goal is to
write a parser of type \code{Parser Program} in terms of lower-level parsers for statements, expressions,
identifiers, etc.

Composing parsing primitives into more complex parsers allows us to avoid an explicit lexing step,
where a stream of characters is broken down into a stream of important tokens like identifiers, string and 
integer literals, punctuation, etc. However, this means your parser needs to explicitly ignore 
whitespace and newlines. Use the \code{token}, \code{strToken}, and \code{whitespace} parsers
to help with this. Try to move whitespace handling to the lowest-level functions so that higher-level
parsers are more readable.

Be sure to consider what happens when your parser finishes parsing a single statement and the remaining
input cannot be parsed. This is probably erroneous. Consider how you might use the \code{eoi} (end-of-input)
parser to require that the program parse the input buffer in its entirety before succeeding.

\subsection{Implementing the interpreter}

Because LANG programs need to read and write to the console while also manipulating the global
name scope, you will need to compose the \code{IO} monad with the \code{StateT} transformer.
You will also need to write a function that can execute an action in this combined monad.

Use the \code{Map} type from the \code{Data.Map} module to map \code{String} identifiers to their
values.

Consider writing a function for interpreting a single statement that uses pattern matching to
consider each of LANG's statement types. How might you implement \code{while}-loop iteration
in Haskell? How can you interpret a \code{Program}, i.e., a list of \code{Statement}s, without
explicit recursion?

Consider writing some basic primitive operations for your composed \code{IO}/\code{StateT} monad
for loading the value of an variable, storing a new value in a variable, reading and writing 
integers, and echoing strings to the console. Write your interpreter function in terms of these
primitives.

\end{document}
