\documentclass[12pt]{article}
% \usepackage{palatino}
\usepackage{epsfig}
\usepackage{epstopdf} % for \psfig{.eps}
\usepackage{amssymb} % for \varnothing
\usepackage{amsmath} % for \pmod
\usepackage{url} % to split long names
\usepackage{color}
\usepackage{listings}
\usepackage{hyperref}
\setlength{\topmargin}{-0.1in}
\setlength{\textheight}{8.0in}
% \newcommand\note[1]{{\color{red} #1}}
\pagestyle{myheadings}
\markboth{Haskell Lecture Notes}{Haskell Lecture Notes}
\newcommand\note[1]{\raggedright\fbox{#1}}
% \newcommand\ind{\hspace*{.3in}}
\newcommand\code[1]{\texttt{\textbf{#1}}}
\def\>{\hspace*{0.2in}}
\newenvironment{notelist}{\begin{list}
   {$\bullet$}
   {\setlength{\itemsep}{0in}}}
   {\end{list}}

\definecolor{light-gray}{gray}{0.75}

% Options for code listings
\lstset{
    aboveskip=\baselineskip,
    basicstyle=\ttfamily\small,
    frame=none,
    language=Haskell,
    numberstyle=\ttfamily\small\color{light-gray},
    numbers=right,
    showstringspaces=false,
}

\setlength{\parindent}{0pt}
\setlength{\parskip}{\baselineskip}

\author{Zachary May}
\title{Haskell Lecture Notes}
\begin{document}
\maketitle

\input{introduction.tex}
\input{type-system.tex}
\input{typeclasses.tex}
\input{functors.tex}
\input{applicative-functors.tex}
\input{monads.tex}
\input{parser-combinators.tex}
\input{monad-transformers.tex}

\begin{thebibliography}{99}
\bibitem{haskell98}
  Simon Peyton Jones, et al.,
  \href{http://www.haskell.org/onlinereport/index.html}{\emph{Haskell 98 Language and Libraries: The Revised Report}},
  2002.
\bibitem{haskellhistory}
  Paul Hudak, et al.,
  \href{http://www.scs.stanford.edu/\~dbg/readings/haskell-history.pdf}{``A History of Haskell: Being Lazy With Class''},
  2007.
\bibitem{hope}
  R.M. Burstall, D.B. MacQueen, D.T. Sannella,
  \href{http://homepages.inf.ed.ac.uk/dts/pub/hope.pdf}{``Hope: An Experimental Applicative Language''},
  1980.
\bibitem{derivabletype}
  Ralf Hinze, Simon Peyton Jones,
  ``Derivable Type Classes'',
  \emph{Proceedings of the Fourth Haskell Workshop}, 227--236,
  2000.
\bibitem{typeclassopedia}
  Brent Yorgey,
  ``The Typeclassopedia'',
  \href{http://www.haskell.org/wikiupload/8/85/TMR-Issue13.pdf}{\emph{The Monad.Reader}}, 13, 17-68,
  2009.
\bibitem{ghc}
  GHC Documentation,
  \href{http://www.haskell.org/ghc/docs/7.8.3/html/},
  2014.
\bibitem{applicative}
  Conor McBride, Ross Paterson,
  ``Applicative Programming with Effects''
  \href{}{\emph{Journal of Functional Programming}} 18:1, 1-13,
  2008.
\bibitem{learnyou}
  Milan Lipova\v{c}a,
  \emph{Learn You a Haskell for Great Good},
  2011.
\bibitem{awkward}
  Simon Peyton Jones,
  \href{http://research.microsoft.com/en-us/um/people/simonpj/papers/marktoberdorf/mark.pdf}
       {``Tackling the Awkward Squad: monadic input/output, concurrency, exceptions, and foreign-language calls in Haskell''},
  2010.
\end{thebibliography}

\end{document}
